SeedLing: Building and using a seed corpus \\ for the Human Language Project

### 2014-04-28 docs/computEL-final/clean.tex
Reworked discussion of experiments, trimmed down to 8 pages.

>>>>

Some languages have more than one orthography. For example, Mandarin Chinese is written with either traditional or simplified characters; Serbian is written with either the Cyrillic or the Roman alphabet. For cross-linguistic NLP, it could be helpful to have standard codes to identify orthographies, but at present none exist.

### 2014-04-28 docs/computEL-final/cluster.tex
Reworked discussion of experiments, trimmed down to 8 pages.

To exemplify the use of SeedLing for computational research on low-resource languages, we experiment with automatic detection of similar languages. When working on endangered languages, documentary and computational linguists alike face the issue of lack of resources. It is often helpful to exploit lexical, syntactic or morphological knowledge about related or similar languages. For example, information about relatedness can be useful for identifying high-resource languages to be used in bootstrapping approaches such as those described in \newcite{yarowsky:ngai:2001} or \newcite{xia2007multilingual}.

@@

We test four methods, which differ in how they define the distance between two clusters. The \texttt{complete} method considers all possible pairs with one language from each cluster, and finds the largest distance. The \texttt{single} method instead finds the smallest distance. The \texttt{average} and \texttt{weighted} methods calculate averages across all the pairwise differences, either unweighted, or weighted according to the languages' positions in the tree.

To ease evaluation, we produce a partitional clustering, by stopping when we reach a certain number of clusters. We set the number of clusters to 147, the number of top-level genetic groupings in Ethnologue.

\begin{table}[t]
\begin{centering}

    \begin{tabular}{l|ccc}
    ~        & Precision & Recall       & F-score    \\ \hline
	\emph{random} & 0.1836	& 0.0917	& 0.0675 \\ \hline
	\texttt{complete} & \textbf{0.3142}	& 0.1938	 & \textbf{0.1556} \\
	\texttt{single} & 0.1949	& \textbf{0.7850}	 & 0.1282 \\
	\texttt{average} & 0.2216	& 0.5163	& 0.1348 \\
	\texttt{weighted} & 0.2326	& 0.3343	& 0.1271 \\
    \end{tabular}
\caption{Comparison of clustering algorithms}
\label{table:cluster}
\end{centering}
\end{table}

\paragraph{Evaluation.}
We compare our induced clusters to the language families given in Ethnologue. However, there are many possible metrics to evaluate the quality of a clustering. \newcite{amigo2009metrics} propose a set of criteria which a clustering evaluation metric should satisfy, and demonstrate that most popular metrics fail to satisfy at least one of these criteria.  However, they prove that all criteria are satisfied by the BCubed metric, which we adopt for this reason.  To calculate the BCubed score, we take the induced cluster and gold standard class for each language, and calculate the F-score of the cluster compared to the class.  These F-scores are then averaged across all languages.

In Table \ref{table:cluster}, we give results of our clustering experiments, comparing the three clustering methods to a random baseline (clustering based on random distances, averaged over 20 runs).  It is worth noting that precision is higher than recall.  This is perhaps to be expected, given that related languages using wildly differing orthographies will appear different.  Nonetheless, our system is reasonably capable of identifying languages which are both related and written similarly.

To allow a closer comparison with \newcite{georgi2010wals}, we can also calculate pairwise scores - i.e. considering when pairs of languages are in the same cluster or the same class. For 147 clusters, the \texttt{complete} method gives a pairwise f-score of 0.147, while Georgi et al. report 0.140. The figures are not quite comparable since we are evaluating over a different set of languages; nonetheless, this result is suggestive, given that we only use surface features, while Georgi et al. use typological features taken from WALS.  This demonstrates that it possible for cross-linguistic research to be conducted even based on extremely shallow features.

%%% [ADD THIS!] As well as performing clustering, we can view this as an information retrieval task: given a language, what are similar languages?  To do this, we found the languages with the closest vectors of n-grams and words.  Since BCubed is calculated averaging across languages and not clusters, we can use exactly the same calculation, using the set of nearby languages in place of a cluster.

\paragraph{Qualitative Inspection.}
Interpreting the clusters directly is difficult, because of a large amount of noise in the data.
Using the \texttt{complete} method with 147 clusters, we obtain a mix of large and small clusters. There are 27 singleton clusters, while the largest cluster contains 123 languages. Although these singleton clusters belong to classes of various size, it is interesting that the distribution of sizes is similar to the distribution in the gold standard data, where we have 27 single classes, and the largest class contains 150 languages.  This isn't true for the other clustering methods - the \texttt{single} method produces 138 singleton clusters, with almost all languages in a single cluster of size 968.

>>>>

To exemplify the use of SeedLing for computational research on low-resource languages, we experiment with automatic detection of similar languages. When working on endangered languages, documentary and computational linguists alike face a lack of resources. It is often helpful to exploit lexical, syntactic or morphological knowledge of related languages. For example, similar high-resource languages can be used in bootstrapping approaches, such as described by \newcite{yarowsky:ngai:2001} or \newcite{xia2007multilingual}.

@@

We define the distance between two clusters by considering all possible pairs of languages, with one from each cluster, and taking the largest distance. We experimented with other ways to define the distance between clusters, but results were poor and we omit results for brevity.

To ease evaluation, we produce a partitional clustering, by stopping when we reach a certain number of clusters, set in advance.

\begin{table}[t]
\centering
    \begin{tabular}{l|ccc}
    ~        & Precision & Recall       & F-score    \\ \hline
	Base. 1: random	& 0.184	& 0.092	& 0.068 \\
	Base. 2: together	& 0.061	& 1.000	& 0.112 \\
	Base. 3: separate	& 1.000	& 0.086	& 0.122 \\ \hline
	SeedLing	& 0.255	& 0.205	& 0.150 
    \end{tabular}
\caption{Clustering compared with baselines}
\label{table:cluster}
\end{table}

\begin{figure}[t]
\begin{centering}
\includegraphics[scale=0.75]{performance.png}
\caption{Performance against number of clusters}
\label{fig:performance}
\end{centering}
\end{figure}

\paragraph{Evaluation.}
We compare our clustering to the language families in Ethnologue. However, there are many ways to evaluate clustering quality. \newcite{amigo2009metrics} propose a set of criteria which a clustering evaluation metric should satisfy, and demonstrate that most popular metrics fail to satisfy at least one of these criteria.  However, they prove that all criteria are satisfied by the BCubed metric, which we therefore adopt.  To calculate the BCubed score, we take the induced cluster and gold standard class for each language, and calculate the F-score of the cluster compared to the class.  These F-scores are then averaged across all languages.

In Table \ref{table:cluster}, we set the number of clusters to be 105, the number of language families in our data, and compare this with three baselines: a random baseline (averaged over 20 runs); putting all languages in a single cluster; and putting all languages in separate clusters. Our clustering outperforms all baselines. It is worth noting that precision is higher than recall, which is perhaps expected, given that related languages using wildly differing orthographies will appear distinct.

To allow a closer comparison with \newcite{georgi2010wals}, we calculate pairwise scores - i.e. considering if pairs of languages are in the same cluster or the same class. For 105 clusters, we achieve a pairwise f-score of 0.147, while Georgi et al. report 0.140. The figures are not quite comparable since we are evaluating over a different set of languages; nonetheless, we only use surface features, while Georgi et al. use typological features from WALS.  This suggests the possibility for cross-linguistic research to be conducted based on shallow features.

In Figure \ref{fig:performance}, we vary the number of clusters. The highest f-score is obtained for 199 clusters. There is a notable jump in performance between 98 and 99, just before the true number of families, 105.

Interpreting the clusters directly is difficult, because they are noisy.  However, the distribution of cluster sizes mirrors the true distribution - for 105 clusters, we have 48 clusters of size 1 or 2, with the largest cluster of size 130; while in fact there are 51 families with only 1 or 2 languages in the data, with the largest of size 150.

### 2014-04-28 docs/computEL-final/concl.tex
Reworked discussion of experiments, trimmed down to 8 pages.

In this paper, we have described the creation of SeedLing, a foundation text for a universal corpus, following the guidelines of Abney and Bird (2010; 2011). To do this, we cleaned and standardised data from several multilingual data sources: ODIN, Omniglot, the UDHR, Wikipedia. The resulting corpus is more easily machine-readable than any of the underlying data sources, and has been stored according to the best practices suggested by \newcite{bird2003port}. At present, SeedLing has data from 19\% of the world's living languages, covering 72\% of language families. We believe that a corpus with such diversity of languages, uniformity of format, cleanliness of data, and ease of access provides an excellent seed for a universal corpus. It is our hope that taking steps toward creating this resource will spur both further data contributions and interesting computational research with cross-linguistic or typological perspectives; we have here demonstrated SeedLing's utility for such research by using the data to perform language clustering, with promising results.

>>>>

In this paper, we have described the creation of SeedLing, a foundation text for a Universal Corpus, following the guidelines of Abney and Bird (2010; 2011). To do this, we cleaned and standardised data from several multilingual data sources: ODIN, Omniglot, the UDHR, Wikipedia. The resulting corpus is more easily machine-readable than any of the underlying data sources, and has been stored according to the best practices suggested by \newcite{bird2003port}. At present, SeedLing has data from 19\% of the world's living languages, covering 72\% of language families. We believe that a corpus with such diversity of languages, uniformity of format, cleanliness of data, and ease of access provides an excellent seed for a universal corpus. It is our hope that taking steps toward creating this resource will spur both further data contributions and interesting computational research with cross-linguistic or typological perspectives; we have here demonstrated SeedLing's utility for such research by using the data to perform language clustering, with promising results.

### 2014-04-28 docs/computEL-final/data.tex
New new heatmap
Update number of languages and families.
Reworked discussion of experiments, trimmed down to 8 pages.

\textbf{Aligned texts} consist of multiple parallel documents, aligned at the document, sentence, or word level. The collection of parallel documents as a whole is assigned a unique identifier, with each individual document labelled by this identifier and the corresponding language code.  A monolingual document is simply one with only a single associated language code.  Sentences and words within documents are also assigned identifiers (relative to the document), if alignment has been performed at that level.

\textbf{Analysed texts}, in addition to the raw text, contain more detailed annotations including parts of speech, morphological information, and syntactic relations.  This is stored as a table, where rows represent words, and columns represent the following types of information: document ID, language code, sentence ID, word ID, wordform, lemma, morphological information, part of speech, gloss, head/governor, and relation/role.

@@

Although data size matters in general NLP, \emph{universality} is the top priority for a universal corpus. We focus on the following data sources, because they include a large number of languages, include several parallel texts, and demonstrate a variety of data types which a linguist might encounter (structured, semi-structured, unstructured): Online Database of Interlinear Text (ODIN), Omniglot website, Universal Declaration of Human Rights (UHDR), and Wikipedia.

@@

\begin{table}[t]
\small
\centering
    \begin{tabular}{l|rr|rr}
    ~         				& Langs. 	& Families 	& Tokens		& Size	\\ \hline
    ODIN      				& 1,270      & 100       		& 351,161		& 39 MB		\\
    Omniglot  				& 129        & 20        		& 31,318		& 677 KB	\\
    UDHR      				& 352        & 46        		& 640,588		& 5.2 MB	\\
    Wikipedia 				& 271        & 21       		&				& 37 GB\\ \hline
    \textbf{Combined}	    & 1,451	    & 105 
    \end{tabular}
\caption{Corpus Coverage}
\label{table:corpus}
\end{table}

@@

According to Ethnologue, there are 7105 living languages, and 147 living language families. Across all our data sources, we manage to cover 1347 languages in 106 families, which represents 19.0\% of the world's languages. To get a better idea of the kinds of languages represented, we give a breakdown according to their EGIDS scores (Expanded Graded Intergenerational Disruption Scale) \cite{lewis2010assessing} in Figure \ref{fig:heatmap}. The values in each cell have been colored according to proportion of languages represented, with green indicating good coverage and red poor. It's interesting to note that vigorous languages (6a) are poorly represented across all data sources, and worse than more endangered categories. In terms of language documentation, vigorous languages are less urgent goals than those in categories 6b and up, but this highlights an unexpected gap in linguistic resources.

\begin{figure*}[t]
\begin{centering}
\includegraphics[scale=0.63]{heatmap-numbers.png}
%\includegraphics{heatmap-numbers.png}
\caption{Heatmap of languages in SeedLing according to endangerment status}\label{fig:heatmap}
\end{centering}
\end{figure*}

>>>>

\textbf{Aligned texts} consist of parallel documents, aligned at the document, sentence, or word level. Note that monolingual documents are viewed as aligned texts only tied to a single language.

\textbf{Analysed texts}, in addition to the raw text, contain more detailed annotations including parts of speech, morphological information, and syntactic relations.  This is stored as a table, where rows represent words, and columns represent: document ID, language code, sentence ID, word ID, wordform, lemma, morphological information, part of speech, gloss, head/governor, and relation/role.

@@

Although data size matters in general NLP, \emph{universality} is the top priority for a universal corpus. We focus on the following data sources, because they include a large number of languages, include several parallel texts, and demonstrate a variety of data types which a linguist might encounter (structured, semi-structured, unstructured): the Online Database of Interlinear Text (ODIN), the Omniglot website, the Universal Declaration of Human Rights (UHDR), and Wikipedia.

@@

\begin{table}[t]
\small
\centering
    \begin{tabular}{l|rr|rr}
    ~         				& Langs. 	& Families 	& Tokens		& Size	\\ \hline
    ODIN      				& 1,270      & 100       		& 351,161		& 39 MB		\\
    Omniglot  				& 129        & 20        		& 31,318		& 677 KB	\\
    UDHR      				& 352        & 46        		& 640,588		& 5.2 MB	\\
    Wikipedia 				& 271        & 21       		&				& 37 GB\\ \hline
    Combined	    & 1,451	    & 105 
    \end{tabular}
\caption{Corpus Coverage}
\label{table:corpus}
\end{table}

\begin{figure*}[t]
\begin{centering}
\includegraphics[scale=0.5]{heatmap2.png}
\caption{Heatmap of languages in SeedLing according to endangerment status}\label{fig:heatmap}
\end{centering}
\end{figure*}

@@

According to Ethnologue, there are 7105 living languages, and 147 living language families. Across all our data sources, we manage to cover 1451 languages in 105 families, which represents 19.0\% of the world's languages. To get a better idea of the kinds of languages represented, we give a breakdown according to their EGIDS scores (Expanded Graded Intergenerational Disruption Scale) \cite{lewis2010assessing} in Figure \ref{fig:heatmap}. The values in each cell have been colored according to proportion of languages represented, with green indicating good coverage and red poor. It's interesting to note that vigorous languages (6a) are poorly represented across all data sources, and worse than more endangered categories. In terms of language documentation, vigorous languages are less urgent goals than those in categories 6b and up, but this highlights an unexpected gap in linguistic resources.

### 2014-04-28 docs/computEL-final/intro.tex
Reworked discussion of experiments, trimmed down to 8 pages.

At the time of writing, 7105 living languages are documented in Ethnologue,\footnote{\url{http://www.ethnologue.com}} but \newcite{simons2011update} calculate that 37\% of the extant languages at the time were at various stages of losing transmisson to new generations of speakers.
Only a fraction of the world's languages are well documented, fewer have machine-readable resources, and fewer again have resources with linguistic annotations \cite{maxwell2006annotation} - so the time to work on compiling these resources is now.

Several years ago, Abney and Bird \shortcite{abney2010universal,abney2011data} posed the challenge of building a Universal Corpus, calling it the Human Language Project. Such a corpus would include data from all of the world's languages, in a consistent structure, facilitating large-scale cross-linguistic processing. The challenge was issued to the computational linguistics community, from the perspective that the language processing, machine learning, and data manipulation and management tools well-known in computational linguistics must be brought to bear on the problems of documentary linguistics, if we are to make any serious progress toward building such a resource. The Universal Corpus as envisioned would facilitate broadly cross-lingual natural language processing (NLP), in particular driving innovation in research addressing NLP for low-resource languages, which in turn drives developments in automated analysis, supporting the work and increasing the efficiency of the language documentation process. 
%\textcolor{blue}{[cite our AL work? is this too strong? also, run-on sentence. also, perhaps mention this workshop, which has precisely these goals]}

We have accepted this challenge and have begun converting existing resources into a format consistent with Abney and Bird's specifications for a universal corpus. We aim for a collection of resources that includes data: (a) from as many languages as possible, and (b) in a format which is both in accordance with best practice archiving recommendations and readily accessible as training data for machine learning and for other computational methods. Of course there are many relevant efforts toward producing cross-linguistic resources, which we survey in section \ref{sec:related}. To the best of our knowledge, though, no existing effort meets these two desiderata to the extent of our corpus, which we name SeedLing: a seed corpus for the Human Language Project.

To produce SeedLing, we have drawn on four web sources, described in section \ref{sec:sources}. To bring the four resources into a single common format and data structure (section \ref{sec:structure}), each required different degrees and types of cleaning and standardisation. We describe the steps required in section \ref{sec:case_studies}, presenting each resource as a separate mini-case study. We hope to make future resource conversion efforts more efficient, as they can be guided by the lessons we learned in assembling our seed corpus. To that end, many of the resources described in section~\ref{sec:related} are candidates for inclusion in the next stage of building a universal corpus.

We believe the resulting corpus, which at present covers 1451 languages from 105 language families, is the first of its kind: large enough and consistent enough to allow broadly multilingual language processing. To test this claim, we use SeedLing in a sample application (section \ref{sec:cluster}): the task of language clustering. With no additional pre-processing required, we extract surface-level features (frequencies of character n-grams and words) for estimating the similarity of two languages. Unlike most previous approaches to the task, we make no use of resources curated for aims of linguistic typology (e.g. values of typological features as in WALS \cite{wals}, Swadesh word lists). Despite our approach being highly dependent on orthography, our clustering performance matches the results obtained by \newcite{georgi2010wals} using typolological features, which demonstrates SeedLing's utility in cross-linguistic research.

% \textcolor{blue}{[START HERE WITH CLEARER DESCRIPTION OF CLUSTERING RESULTS]}
% Despite this approach being highly dependent on orthography, we are able to reconstruct substantial parts of several language families, demonstrating the utility of this resource in cross-linguistic research. Finally, we discuss future work in section \ref{sec:future}, and conclude in section \ref{sec:conclusion}.

% Abney and Bird \shortcite{abney2010universal,abney2011data} posed the grand challenge of building a Universal Corpus, calling it the Human Language Project. Such a corpus would include all of the world's languages, in a consistent structure, facilitating large-scale cross-linguistic processing. They propose a specific data structure, which we describe and discuss in section \ref{sec:structure}. We have accepted their challenge, and have begun converting existing resources into a format consistent with their specifications for a universal corpus. We have drawn on four [or five?] web sources, cleaning and standardising them as described in section \ref{sec:sources}, to produce a seed corpus for the Human Language Project. In sections \ref{sec:stats} and \ref{sec:copyright}, we respectively give a summary of the data contained, and discuss copyright and distribution. \textcolor{blue}{[needs rewriting following reorganization.]}

% [Note: remove this section if we don't do the work!] We believe the resulting corpus is the first of its kind: large enough and consistent enough to allow language processing on a grand scale. In section \ref{sec:cluster}, we give an example application of this corpus: language clustering. We use the frequencies of character n-grams and words to estimate the similarity of two languages.

>>>>

At the time of writing, 7105 living languages are documented in Ethnologue,\footnote{\url{http://www.ethnologue.com}} but \newcite{simons2011update} calculated that 37\% of extant languages were at various stages of losing transmisson to new generations. Only a fraction of the world's languages are well documented, fewer have machine-readable resources, and fewer again have resources with linguistic annotations \cite{maxwell2006annotation} - so the time to work on compiling these resources is now.

Several years ago, Abney and Bird \shortcite{abney2010universal,abney2011data} posed the challenge of building a Universal Corpus, naming it the Human Language Project. Such a corpus would include data from all the world's languages, in a consistent structure, facilitating large-scale cross-linguistic processing. The challenge was issued to the computational linguistics community, from the perspective that the language processing, machine learning, and data manipulation and management tools well-known in computational linguistics must be brought to bear on the problems of documentary linguistics, if we are to make any serious progress toward building such a resource. The Universal Corpus as envisioned would facilitate broadly cross-lingual natural language processing (NLP), in particular driving innovation in research addressing NLP for low-resource languages, which in turn supports the language documentation process. 
%\textcolor{blue}{[cite our AL work? is this too strong? also, run-on sentence. also, perhaps mention this workshop, which has precisely these goals]}

We have accepted this challenge and have begun converting existing resources into a format consistent with Abney and Bird's specifications. We aim for a collection of resources that includes data: (a) from as many languages as possible, and (b) in a format both in accordance with best practice archiving recommendations and also readily accessible for computational methods. Of course, there are many relevant efforts toward producing cross-linguistic resources, which we survey in section \ref{sec:related}. To the best of our knowledge, though, no existing effort meets these two desiderata to the extent of our corpus, which we name SeedLing: a seed corpus for the Human Language Project.

To produce SeedLing, we have drawn on four web sources, described in section \ref{sec:sources}. To bring the four resources into a single common format and data structure (section \ref{sec:structure}), each required different degrees and types of cleaning and standardisation. We describe the steps required in section \ref{sec:case_studies}, presenting each resource as a separate mini-case study. We hope that the lessons we learned in assembling our seed corpus can guide future resource conversion efforts. To that end, many of the resources described in section~\ref{sec:related} are candidates for inclusion in the next stage of building a Universal Corpus.

We believe the resulting corpus, which at present covers 1451 languages from 105 language families, is the first of its kind: large enough and consistent enough to allow broadly multilingual language processing. To test this claim, we use SeedLing in a sample application (section \ref{sec:cluster}): the task of language clustering. With no additional pre-processing, we extract surface-level features (frequencies of character n-grams and words) to estimate the similarity of two languages. Unlike most previous approaches to the task, we make no use of resources curated for linguistic typology (e.g. values of typological features as in WALS \cite{wals}, Swadesh word lists). Despite our approach being highly dependent on orthography, our clustering performance matches the results obtained by \newcite{georgi2010wals} using typolological features, which demonstrates SeedLing's utility in cross-linguistic research.

### 2014-04-28 docs/computEL-final/relwork.tex
Reworked discussion of experiments, trimmed down to 8 pages.

In this section, we review existing efforts to compile multilingual machine-readable resources.  Although some commercial resources are available, we restrict attention to freely accessible data.\footnote{All figures given below were correct at the time of writing, but it must be borne in mind that most resources discussed below are constantly growing.}

\paragraph{Traditional archives.}
Many archives exist to store the wealth of traditional resources produced by the documentary linguistics community.  Such documents are increasingly being digitised, or produced in a digital form, and there are a number of archives which now offer free online access to their data.

Some archives aim for a universal scope, including languages from all parts of the world.  Of particular note are: The Language Archive, maintained by the Max Planck Institute of Psycholinguistics, which includes DoBeS (Dokumentation Bedrohter Sprachen) and many other projects;
Collection Pangloss, maintained by LACITO (Langues et Civilisations à Tradition Orale);
The Endangered Languages Archive (ELAR), maintained by the School of Oriental and African Studies, which forms one part of the Hans Rausing Endangered Languages Project (HRELP).

Regional archives include: AILLA (Archive of Indigenous Languages of Latin America), University of Texas at Austin;
AIATSIS collections (Australian Institute of Aboriginal and Torres Strait Islander Studies);
California Language Archive, managed by the Survey of California and Other Indian Languages;
ANLA (Alaska Native Language Archive), at the University of Alaska Fairbanks;
PARADISEC (Pacific and Regional Archive for Digital Sources in Endangered Cultures).

>>>>

In this section, we review existing efforts to compile multilingual machine-readable resources.  Although some commercial resources are available, we restrict attention to freely accessible data.\footnote{All figures given below were correct at the time of writing, but it must be borne in mind that most of these resources are constantly growing.}

\paragraph{Traditional archives.}
Many archives exist to store the wealth of traditional resources produced by the documentary linguistics community.  Such documents are increasingly being digitised, or produced in a digital form, and there are a number of archives which now offer free online access.

Some archives aim for a universal scope, such as The Language Archive (maintained by the Max Planck Institute of Psycholinguistics), Collection Pangloss (maintained by LACITO), and The Endangered Languages Archive (maintained by SOAS). Most archives are regional, including AILLA, ANLA, PARADISEC, and many others.
